\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2021-06-13 Papers/Stochastic_Uncertainty_Propagation_TPS/JKO_Power_Systems_TPS.tex
updated TPS manuscript

We mention here that (\ref{ItoSDEvectorlevel}) has been used in \cite{ju2018analytical,ju2018stochastic} for uncertainty propagation via stochastic averaging approximation where the univariate energy PDF was proposed as a ``proxy" for the entire joint PDF. Most relevant to our approach in the power systems literature is the work in \cite{wang2013fokker}, which indeed voiced the need for computing the transient joint PDFs but only dealt with the single-machine-infinite-bus case -- simplest instance of (\ref{ItoSDEvectorlevel}). The resulting bivariate Fokker-Planck PDE in \cite{wang2013fokker} was solved via finite element discretization, and revealed rich stochastic dynamics and nontrivial transient stability aspects even in this simple case. However, it is unreasonable to expect that a finite element discretization, or in fact any spatial discretization scheme to solve (\ref{KineticFPK}) for moderately large $n$ in minutes of computational time, thereby limiting our current ability for realistic power systems simulation with stochastic variability. This calls for fundamentally re-thinking what does it mean to solve the PDE (\ref{KineticFPK}) for dynamics (\ref{ItoSDEvectorlevel}).

@@

\subsection{Statistical Mechanics Perspective}
A new difficulty for our SDE \eqref{ItoSDEvectorlevel} is that we have \emph{anisotropic} degenerate diffusion, i.e., the strengths of the noise acting in the last $n$ components of \eqref{ItoSDEvectorlevel} are nonuniform since $\bm{M}^{-1}\bm{\Sigma}$ is not identity. This complicates the matter because the construction of the functional $\Phi$ in \eqref{ProxRecursionInfiniteDim} is usually motivated via free energy considerations utilizing the structure of the \emph{stationary} PDF $\rho_{\infty}(\bm{\theta},\bm{\omega})$ for \eqref{KineticFPK}. The $\rho_{\infty}$ is, in turn, guaranteed to be a unique \emph{Boltzmann distribution} of the form
\begin{subequations}
\begin{align}
\rho_{\infty}\left(\bm{\theta},\bm{\omega}\right) &= \dfrac{1}{Z}\exp\left(-\beta H\right), \\
H(\bm{\theta},\bm{\omega}) &:= V(\bm{\theta}) + \frac{1}{2}\langle\bm{\omega},\bm{M\omega}\rangle,	
\end{align}	
\label{BoltzmannPDF}
\end{subequations}
where $Z$ is a normalizing constant (known as the ``partition function"), \emph{if and only if} the so-called \emph{Einstein relation} \cite{hernandez1989equilibrium,chen2015fast} holds:
\begin{align}
\bm{\Sigma}\bm{\Sigma}^{\top} = \beta^{-1}\left(\bm{\Gamma} + \bm{\Gamma}^{\top}\right) \quad\text{for some}\;\beta > 0.
\label{EinsteinRelation}	
\end{align}
In our case, $\bm{\Sigma,\Gamma}$ are positive diagonal, and \eqref{EinsteinRelation} is equivalent to $\sigma_{i}^{2} \propto \gamma_{i}$ for all $i=1,\hdots,n$.

@@

We notice that \eqref{XiEtaVectorSDE} is a mixed-conservative SDE with \emph{isotropic} degenerate diffusion. In particular, the pushforward of the known initial joint PDF \eqref{FPKInitCond} via $\bm{\Psi}$, is given by
\begin{align}
\tilde{\rho}_{0}(\bm{\xi},\bm{\eta}) := \bm{\Psi}_{\sharp} \rho_{0} 
&= \dfrac{\rho_{0}\left(\bm{\Psi}^{-1} \begin{pmatrix}
\bm{\xi}\\
\bm{\eta}	
\end{pmatrix}\right)}{|\det\left(\bm{\Psi}\right)|} \nonumber\\
&= \dfrac{\rho_{0}\left(\bm{\Sigma}\bm{M}^{-1}\bm{\xi},\bm{\Sigma}\bm{M}^{-1}\bm{\eta}\right)}{\left(\displaystyle\prod_{i=1}^{n}m_{i}/\sigma_{i}\right)^{\!2}},
\label{PushforwardOfIC}	
\end{align}
where we used the standard properties of the Kronecker product. The transient joint state PDF $\tilde{\rho}(t,\bm{\xi},\bm{\eta})$ corresponding to \eqref{XiEtaVectorSDE} solves the PDE initial value problem
\begin{subequations}
\begin{align}
&\dfrac{\partial\tilde{\rho}}{\partial t} = -\langle\bm{\eta},\nabla_{\bm{\xi}}\tilde{\rho}\rangle + \nabla_{\bm{\eta}}\cdot\left(\tilde{\rho}\left(\nabla_{\bm{
\xi}}U(\bm{\xi})+\nabla_{\bm{\eta}}F(\bm{\eta})\right)\right) + \dfrac{1}{2}\Delta_{\bm{\eta}}\tilde{\rho}, \label{NewPDE}\\
&\tilde{\rho}(t=t_{0},\bm{\xi},\bm{\eta}) = \underbrace{\tilde{\rho}_{0}(\bm{\xi},\bm{\eta})}_{\text{from}\;\eqref{PushforwardIC}}. \label{NewIC}	
\end{align}
\label{TransformedPDEivp}	
\end{subequations}
In other words, \eqref{TransformedPDEivp} is the macroscopic dynamics corresponding to the sample path dynamics \eqref{XiEtaVectorSDE}.

Since \eqref{NewPDE} is a standard kinetic Fokker-Planck PDE with isotropic degnerate diffusion, our strategy is to perform a proximal recursion of the form \eqref{ProxRecursionInfiniteDim} for \eqref{TransformedPDEivp}, and then to pushforward the resulting joint PDFs via $\bm{\Psi}^{-1}$ to the original $(\bm{\theta},\bm{\omega})$ state space.

\subsection{Proximal Update}\label{subsec:ProxUpdate}

>>>>

We mention here that (\ref{ItoSDEvectorlevel}) has been used in \cite{ju2018analytical,ju2018stochastic} for uncertainty propagation via stochastic averaging approximation where the univariate energy PDF was proposed as a ``proxy" for the entire joint PDF. Most relevant to our approach in the power systems literature is the work in \cite{wang2013fokker}, which indeed voiced the need for computing the transient joint PDFs but only dealt with the single-machine-infinite-bus case -- simplest instance of (\ref{ItoSDEvectorlevel}). The resulting bivariate Fokker-Planck PDE in \cite{wang2013fokker} was solved via finite element discretization, and revealed rich stochastic dynamics and nontrivial transient stability aspects even in this simple case. However, it is unreasonable to expect that a finite element discretization, or in fact any spatial discretization scheme to solve (\ref{KineticFPK}) for moderately large $n$ in {\red{minutes}} of computational time, thereby limiting our current ability for realistic power systems simulation with stochastic variability. This calls for fundamentally re-thinking what does it mean to solve the PDE (\ref{KineticFPK}) for dynamics (\ref{ItoSDEvectorlevel}).

@@

\subsection{Statistical Mechanics Perspective}
A new difficulty for our SDE \eqref{ItoSDEvectorlevel} is that we have \emph{anisotropic} degenerate diffusion, i.e., the strengths of the noise acting in the last $n$ components of \eqref{ItoSDEvectorlevel} are nonuniform since $\bm{M}^{-1}\bm{\Sigma}$ is not identity. This complicates the matter because the construction of the functional $\Phi$ in \eqref{ProxRecursionInfiniteDim} is usually motivated via free energy considerations utilizing the structure of the \emph{stationary} PDF $\rho_{\infty}(\bm{\theta},\bm{\omega})$ for \eqref{KineticFPK}. The $\rho_{\infty}$ is, in turn, guaranteed to be a unique \emph{Boltzmann distribution} of the form\footnote{here $Z$ is a normalizing constant known as the ``partition function".}
\begin{subequations}
\begin{align}
\rho_{\infty}\left(\bm{\theta},\bm{\omega}\right) &= \dfrac{1}{Z}\exp\left(-\beta H\right), \\
H(\bm{\theta},\bm{\omega}) &:= V(\bm{\theta}) + \frac{1}{2}\langle\bm{\omega},\bm{M\omega}\rangle,	
\end{align}	
\label{BoltzmannPDF}
\end{subequations}
\emph{if and only if} the so-called \emph{Einstein relation} \cite{hernandez1989equilibrium,chen2015fast} holds:
\begin{align}
\bm{\Sigma}\bm{\Sigma}^{\top} = \beta^{-1}\left(\bm{\Gamma} + \bm{\Gamma}^{\top}\right) \quad\text{for some}\;\beta > 0.
\label{EinsteinRelation}	
\end{align}
In our case, $\bm{\Sigma,\Gamma}$ are positive diagonal, and \eqref{EinsteinRelation} is equivalent to $\sigma_{i}^{2} \propto \gamma_{i}$ for all $i=1,\hdots,n$.

@@

We notice that \eqref{XiEtaVectorSDE} is a mixed conservative-dissipative SDE with \emph{isotropic} degenerate diffusion. In particular, the pushforward of the known initial joint PDF \eqref{FPKInitCond} via $\bm{\Psi}$, is given by
\begin{align}
\tilde{\rho}_{0}(\bm{\xi},\bm{\eta}) := \bm{\Psi}_{\sharp} \rho_{0} 
&= \dfrac{\rho_{0}\left(\bm{\Psi}^{-1} \begin{pmatrix}
\bm{\xi}\\
\bm{\eta}	
\end{pmatrix}\right)}{|\det\left(\bm{\Psi}\right)|} \nonumber\\
&= \dfrac{\rho_{0}\left(\bm{\Sigma}\bm{M}^{-1}\bm{\xi},\bm{\Sigma}\bm{M}^{-1}\bm{\eta}\right)}{\left(\displaystyle\prod_{i=1}^{n}m_{i}/\sigma_{i}\right)^{\!2}},
\label{PushforwardOfIC}	
\end{align}
where we used the standard properties of the Kronecker product. The transient joint state PDF $\tilde{\rho}(t,\bm{\xi},\bm{\eta})$ corresponding to \eqref{XiEtaVectorSDE} solves the PDE initial value problem
\begin{subequations}
\begin{align}
&\dfrac{\partial\tilde{\rho}}{\partial t} = -\langle\bm{\eta},\nabla_{\bm{\xi}}\tilde{\rho}\rangle + \nabla_{\bm{\eta}}\cdot\left(\tilde{\rho}\left(\nabla_{\bm{
\xi}}U(\bm{\xi})+\nabla_{\bm{\eta}}F(\bm{\eta})\right)\right) + \dfrac{1}{2}\Delta_{\bm{\eta}}\tilde{\rho}, \label{NewPDE}\\
&\tilde{\rho}(t=t_{0},\bm{\xi},\bm{\eta}) = \underbrace{\tilde{\rho}_{0}(\bm{\xi},\bm{\eta})}_{\text{from}\;\eqref{PushforwardOfIC}}. \label{NewIC}	
\end{align}
\label{TransformedPDEivp}	
\end{subequations}
In other words, \eqref{TransformedPDEivp} is the macroscopic dynamics corresponding to the sample path dynamics \eqref{XiEtaVectorSDE}.

Since \eqref{NewPDE} is a kinetic Fokker-Planck PDE with isotropic degenerate diffusion, our strategy is to perform a proximal recursion of the form \eqref{ProxRecursionInfiniteDim} for \eqref{TransformedPDEivp}, and then to pushforward the resulting joint PDFs via $\bm{\Psi}^{-1}$ to the original state space. This is what we detail next.

\subsection{Proximal Update}\label{subsec:ProxUpdate}
Looking at \eqref{NewPotentials} and \eqref{NewPDE}, it is natural to consider the energy functional 
\begin{align}
\Phi(\tilde{\rho}) := \displaystyle\int_{\mathbb{T}^{n}\times\mathbb{R}^{n}}\left(U(\bm{\xi})+F\left(\bm{\eta}\right) + \frac{1}{2}\log\tilde{\rho}\right)\tilde{\rho}\:\differential\bm{\xi}\differential\bm{\eta},
\label{LyapButNotProx}	
\end{align}
which is the sum of a potential energy (expected value of $U$), a kinetic energy (expected value of $F$), and an internal energy (scaled negative entropy). Indeed, the functional \eqref{LyapButNotProx} is a Lyapunov functional along the solution of \eqref{NewPDE}. To see this, ...

However, it

\section{Proximal Algorithm}\label{sec:ProxAlgo}

\section{Numerical Simulations}\label{sec:NumericalSimulations}
